A search algorithm for the global extremum \break of a discontinuous function\thanks{This work was supported by the Russian Science Foundation, project No.\,21-11-00204.}

### 2021-09-24 2021/ICCS_BarkalovUsova/iccs.tex
answers in Russian

%%%% ДЛЯ РУССКОГО ТЕКСТА закомментировать потом!
%\usepackage{inputenc}
%\usepackage[english,russian]{babel}
%\usepackage{cmap}
%%%%

@@

Global optimization problems involving many local extrema arise in many areas of applications (see, e.g., \cite{Cavoretto2021,Kvasov2015,Kvasov2013,Modorskii2016}). As a rule, the calculation of even a single value of the objective function in such problems requires considerable time, since the analytical form of the function is not known and the calculation of its values involves time-consuming numerical modelling. In this case, we have to deal with expensive black-box optimization problems as they are often called. Problems of this class are extremely complicated and are the object of intensive research.

One should note an important role of one-dimensional global optimization problems. Firstly, as has already been mentioned, they arise in many applications (see, e.g., \cite{Gillard2017,Sergeyev2020}). Secondly, many approaches to solving multidimensional optimization problems are, in one way or another, based on reducing the solution of the original multidimensional problem to solving a series of related one-dimensional optimization problems (see, for example, popular approaches in \cite{Pinter1996,PaulaviciusZilinskas2014,Sergeyev2017,Sergeyev2013}).

@@

\item For each interval $(x_{i-1},x_i), \; 1\leq i \leq m,$ calculate the characteristic
%\begin{equation}\label{R_i}
\[
\begin{gathered}
R(i)=(1+\gamma_i+\gamma_{i-1} )\Delta_i+(1-\gamma_i-\gamma_{i-1} ) \frac{ (z_i-z_{i-1} )^2}{(r\mu)^2 \Delta_i }-\\
-\frac{2[(1-\gamma_i )(1+\gamma_{i-1} ) z_i+(1+\gamma_i )(1-\gamma_{i-1} ) z_{i-1} ]}{r\mu},
\end{gathered}
%\end{equation}
\]
where $r>1$ is the method parameter.

@@

Note that the erroneous assignment of some discontinuity points from the series (\ref{points}) will not lead to the loss of convergence of the method. The algorithm will interpret such points as a recoverable discontinuity case since the left and right limits of the objective function at such points coincide.

@@

\begin{equation}\label{mu_cond}
\mu(1) \geq \mu(2) \geq ... \geq \mu(m)
\end{equation}
and define the minimum number $p$ satisfying the conditions
\begin{equation}\label{mu_main_cond}
 \frac{\mu(p)}{\mu(p+1)} \geq Q, \; 1 \leq p < q(m-2(s+1)),
\end{equation}
where $Q>1$ and $0<q<1$ are set numbers.

@@

>>>>

%%%% ДЛЯ РУССКОГО ТЕКСТА закомментировать потом!
\usepackage{inputenc}
\usepackage[english,russian]{babel}
\usepackage{cmap}
%%%%

@@

Global optimization problems involving many local extrema arise in many areas of applications (see, e.g., \cite{Cavoretto2021,Kvasov2013,Kvasov2015,Modorskii2016}). As a rule, the calculation of even a single value of the objective function in such problems requires considerable time, since the analytical form of the function is not known and the calculation of its values involves time-consuming numerical modelling. In this case, we have to deal with expensive black-box optimization problems as they are often called. Problems of this class are extremely complicated and are the object of intensive research.

One should note an important role of one-dimensional global optimization problems. Firstly, as has already been mentioned, they arise in many applications (see, e.g., \cite{Gillard2017,Sergeyev2020}). Secondly, many approaches to solving multidimensional optimization problems are, in one way or another, based on reducing the solution of the original multidimensional problem to solving a series of related one-dimensional optimization problems (see, for example, popular approaches in \cite{PaulaviciusZilinskas2014,Pinter1996,Sergeyev2017,Sergeyev2013}).

@@

\item For each interval $(x_{i-1},x_i), \; 1\leq i \leq m,$ calculate the characteristic
\begin{equation}\label{R_i}
\begin{gathered}
R(i)=(1+\gamma_i+\gamma_{i-1} )\Delta_i+(1-\gamma_i-\gamma_{i-1} ) \frac{ (z_i-z_{i-1} )^2}{(r\mu)^2 \Delta_i }-\\
-\frac{2[(1-\gamma_i )(1+\gamma_{i-1} ) z_i+(1+\gamma_i )(1-\gamma_{i-1} ) z_{i-1} ]}{r\mu},
\end{gathered}
\end{equation}
where $r>1$ is the method parameter.

@@

%Russian

\textbf{Remark 1}. Еще раз отметим, что в соответствии с (\ref{gamma}) значение $\gamma_i+\gamma_{i-1}$ может быть равно либо 0 (обе граничных точки интервала принадлежат области непрерывности функции), либо 1 (одна из граничных точек является точкой разрыва). В этом случае формула (\ref{R_i}) иметь вид
\[
R(i)=\Delta_i+\frac{ (z_i-z_{i-1} )^2}{(r\mu)^2 \Delta_i }-\frac{2(z_i + z_{i-1})}{r\mu},
\]
в случае отстутсвия точке разрыва и 
\[
R(i)=2\Delta_i - \frac{4 z_i }{r\mu}, \; R(i)=2\Delta_i - \frac{4 z_{i-1} }{r\mu},
\]
в случае разрыва в точке $x_{i-1}$ или $x_{i}$ соответственно.

\textbf{Remark 2}.
Приведенные выше формулы для вычисления характеристик интервалов могут быть использованы (в модифицированной форме) в алгоритме решения задач невыпуклыми ограничениями \cite{Sergeyev1995,Sergeyev2006_1,Sergeyev2007}. Здесь решение задачи с частично определенными ограничениями неявно сводится к минимизации разрывной функции, в которой точками разрыва являются границы допустимого множества решений задачи.

\textbf{Remark 3}. The erroneous assignment of some discontinuity points from the series (\ref{points}) will not lead to the loss of convergence of the method. The algorithm will interpret such points as a recoverable discontinuity case since the left and right limits of the objective function at such points coincide.

@@

\begin{equation}\label{mu_cond}
\mu(1) \geq \mu(2) \geq ... \geq \mu(m)
\end{equation}
and define the minimum number $p$ satisfying the conditions
\begin{equation}\label{mu_main_cond}
 \frac{\mu(p)}{\mu(p+1)} \geq Q, \; 1 \leq p < q(m-2(s+1)),
\end{equation}
where $Q>1$ and $0<q<1$ are parameters.
%Russian
Значение параметра $Q$ определяет наши априорные предположения о том, во сколько раз относительная разность $\mu_i$, вычисленная на интервале $(x_{i-1},x_i)$, содержащем точку разрыва, будет превышать максимальную разность $\mu$, найденную по всем интервалам, принадлежащим области непрерывности функции. Данное значение задается исследователем, и в проведенных эксмериментах оно было выбрано равным 3.

@@

\textbf{Remark}. В соответствии с (\ref{gamma}) and (\ref{delta_cond}) значение $\gamma_i+\gamma_{i-1}+|\delta_i |$ может быть равно 0 либо 1, в зависимости от наличия либо отсутствия точки разрыва в интервале $(x_{i-1},x_i)$.
Так, если данный интервал не содержит точек разрыва, т.е. $\gamma_i+\gamma_{i-1}+|\delta_i | = 0$, его характеристика (\ref{R_i}) пример вид 
\[
R(i)=\Delta_i+\frac{ (z_i-z_{i-1} )^2}{(r\mu)^2 \Delta_i }-\frac{2(z_i + z_{i-1})}{r\mu}.
\]
Если одна из граничных точек интервала является заданной точкой разрыва, т.е. $\gamma_{i-1} = 1$ или $\gamma_i = 1$, то мы получим соответственно характеристики
\begin{equation}\label{RR}
R(i)=2\Delta_i - \frac{4 z_i }{r\mu}, \; R(i)=2\Delta_i - \frac{4 z_{i-1} }{r\mu}.
\end{equation}
Аналогичные формулы для характеристик интервалов будут получаться, если $(x_{i-1},x_i)$ является интервалом, который содержит незаданную точку разрыва, т.е. когда $|\delta_i | = 1 $. В этом случае мы таже будем получать формулы (\ref{RR}) для $\delta_i = -1$ и $\delta_i = 1$ соответственно.
